---
title: "Using `tidyr` to reshape wide to long data - a success story"
author: "Corey Sparks, Ph.D."
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



When using longitudial survey data, we often have to reshape from long to wide form routinely. For the longest time I'd used `stats::reshape()` in the base R, but I've wanted to try to put this into the tidyverse for a long time, mainly to make use of all the tidyverse verbs.

The problem was that I've found the examples of how to use `pivot_longer()` a bit limited, usually only reshaping one variable into the long form. In my data and examples for my classes, I often have several time-varying variables to reshape, not just 1 or 2. 

Below, I use data from the [Early Childhood Longitudinal Survey - Kindergarten Cohort, 2011](https://nces.ed.gov/ecls/kindergarten2011.asp), which is a large, nationally representative survey data collection project which follows over 18,000 children from kindergarten through 5th grade. 

```{r}
library (car)
library(dplyr)
```


## Data and recodes
First we load our data
```{r, message=FALSE}
load("~/OneDrive - University of Texas at San Antonio//classes/dem7283/class_21_7283/data/eclsk_2011/eclsk_k5.Rdata")
names(eclskk5)<-tolower(names(eclskk5))
#get out only the variables I'm going to use for this example

#subset the data

eclsk.sub<-eclskk5%>%
  select(childid, x_chsex_r, x1locale, x_raceth_r, x2povty, x12par1ed_i, p1curmar,
         x1htotal, x1mscalk5, x2mscalk5, x3mscalk5, x4mscalk5, x5mscalk5, p1hscale,
         p2hscale, p4hscale, x2fsstat2, x4fsstat2, x4fsstat2, x12sesl, x4sesl_i,
         p2parct1, p2parct2, s1_id, p2safepl, x2krceth, p1o2near, x_distpov, w1c0,
         w1p0, w2p0, w1c0str, w1p0str, w4c4p_40, w4c4p_4str,w4c4p_4psu, w1c0psu,
         w1p0psu, x1height, x2height, x4height, x4height, x5height, x1kage_r,
         x2kage_r, x3age, x4age, x5age)


```


 
## Time constant variables

First, I do some recoding of variables. First, we code time **invariant** variables, meaning their values do not change at each wave. 

```{r}
#Non time varying variables
#First we recode some Child characteristics
#Child's sex: recode as male =1
eclsk.sub$male<-Recode(eclsk.sub$x_chsex_r, recodes="1=1; 2=0; -9=NA")

#Recode race with white, non Hispanic as reference using dummy vars
eclsk.sub$hisp<-Recode (eclsk.sub$x_raceth_r, recodes="3:4=1;-9=NA; else=0")
eclsk.sub$black<-Recode (eclsk.sub$x_raceth_r, recodes="2=1;-9=NA; else=0")
eclsk.sub$asian<-Recode (eclsk.sub$x_raceth_r, recodes="5=1;-9=NA; else=0")
eclsk.sub$nahn<-Recode (eclsk.sub$x_raceth_r, recodes="6:7=1;-9=NA; else=0")
eclsk.sub$other<-Recode (eclsk.sub$x_raceth_r, recodes="8=1;-9=NA; else=0")


#Then we recode some parent/mother characteristics
#Mother's education, recode as 2 dummys with HS = reference
eclsk.sub$lths<-Recode(eclsk.sub$x12par1ed_i, recodes = "0:2=1; 3:8=0; else = NA")
eclsk.sub$gths<-Recode(eclsk.sub$x12par1ed_i, recodes = "1:3=0; 4:8=1; else =NA") 

#marital status, recode as 2 dummys, ref= married
eclsk.sub$single<-Recode(eclsk.sub$p1curmar, recodes="4=1; -7:-9=NA; else=0")
eclsk.sub$notmar<-Recode(eclsk.sub$p1curmar, recodes="2:3=1; -7:-9=NA; else=0")


#Then we do some household level variables

#Urban school location = 1
eclsk.sub$urban<-Recode(eclsk.sub$x1locale, recodes = "1:3=1; 4=0; -1:-9=NA")

#poverty level in poverty = 1
eclsk.sub$pov<-Recode(eclsk.sub$x2povty , recodes ="1:2=1; 3=0; -9=NA")

#Household size
eclsk.sub$hhsize<-eclsk.sub$x1htotal

#school % minority student body
eclsk.sub$minorsch<-ifelse(eclsk.sub$x2krceth <0, NA, eclsk.sub$x2krceth/10)

#Unsafe neighborhood
eclsk.sub$unsafe<-Recode(eclsk.sub$p2safepl , recodes = "1:2='unsafe'; 3='safe'; else=NA",as.factor = T)

#school district poverty
eclsk.sub$dist_pov<-ifelse(eclsk.sub$x_distpov==-9, NA, scale(eclsk.sub$x_distpov))
```

## Time varying variables
I have to make the repeated measures of each of my longitudinal variables. These are referred to as **time varying** variables, meaning their values change at each wave.

```{r}
#Longitudinal variables
#recode the outcomes, the  first is the child's math standardized test score in Kindergarten
eclsk.sub$math_1<-ifelse(eclsk.sub$x1mscalk5<0, NA, eclsk.sub$x1mscalk5)
eclsk.sub$math_2<-ifelse(eclsk.sub$x2mscalk5<0, NA, eclsk.sub$x2mscalk5)
eclsk.sub$math_4<-ifelse(eclsk.sub$x4mscalk5<0, NA, eclsk.sub$x4mscalk5)

#Second outcome is child's height for age, continuous outcome
eclsk.sub$height_1<-ifelse(eclsk.sub$x1height<=-7, NA, eclsk.sub$x1height)
eclsk.sub$height_2<-ifelse(eclsk.sub$x2height<=-7, NA, eclsk.sub$x2height)
eclsk.sub$height_4<-ifelse(eclsk.sub$x4height<=-7, NA, eclsk.sub$x4height)

#Age at each wave
eclsk.sub$ageyrs_1<-ifelse(eclsk.sub$x1kage_r<0, NA, eclsk.sub$x1kage_r/12)
eclsk.sub$ageyrs_2<-ifelse(eclsk.sub$x2kage_r<0, NA, eclsk.sub$x2kage_r/12)
eclsk.sub$ageyrs_4<-ifelse(eclsk.sub$x4age<0, NA, eclsk.sub$x4age/12)
eclsk.sub<- eclsk.sub[is.na(eclsk.sub$ageyrs_1)==F, ]

#Height for age z score standardized by sex and age
eclsk.sub$heightz_1<-ave(eclsk.sub$height_1, as.factor(paste(round(eclsk.sub$ageyrs_1, 1.5), eclsk.sub$male)), FUN=scale)
eclsk.sub$heightz_2<-ave(eclsk.sub$height_2, as.factor(paste(round(eclsk.sub$ageyrs_2, 1.5), eclsk.sub$male)), FUN=scale)
eclsk.sub$heightz_4<-ave(eclsk.sub$height_4, as.factor(paste(round(eclsk.sub$ageyrs_4, 1.5), eclsk.sub$male)), FUN=scale)


#Household food insecurity, dichotomous outcome
eclsk.sub$foodinsec_1<-Recode(eclsk.sub$x2fsstat2, recodes="2:3=1; 1=0; else=NA")
eclsk.sub$foodinsec_2<-Recode(eclsk.sub$x2fsstat2, recodes="2:3=1; 1=0; else=NA")
eclsk.sub$foodinsec_4<-Recode(eclsk.sub$x4fsstat2, recodes="2:3=1; 1=0; else=NA")


#Child health assessment Excellent to poor , ordinal outcome
eclsk.sub$chhealth_1<-ifelse(eclsk.sub$p1hscale<0, NA, eclsk.sub$p1hscale)
eclsk.sub$chhealth_2<-ifelse(eclsk.sub$p2hscale<0, NA, eclsk.sub$p2hscale)
eclsk.sub$chhealth_4<-ifelse(eclsk.sub$p4hscale<0, NA, eclsk.sub$p4hscale)

#SES
eclsk.sub$hhses_1<-ifelse(eclsk.sub$x12sesl==-9, NA, scale(eclsk.sub$x12sesl))
eclsk.sub$hhses_2<-ifelse(eclsk.sub$x12sesl==-9, NA, scale(eclsk.sub$x12sesl))
eclsk.sub$hhses_4<-ifelse(eclsk.sub$x4sesl_i==-9, NA, scale(eclsk.sub$x4sesl_i))
```

## Reshaping data into longitudinal format

To analyze data longitudinally, we must reshape the data from its current "wide" format, where each repeated measure is a column, into the "long" format, where there is a single column for a particular variable, and we account for the repeated measurements of each person. In this case, I'm going to use three waves of data, so each child can contribute up to three lines to the data.

The `reshape()` function will do this for us, but below I use a tidy method, using a `pivot_longer()` in the `tidyr` packages. 

For the longest time I'd used `stats::reshape()` in the base R, but I've wanted to try to put this into the tidyverse for a long time. The problem was that I've found the examples of how to use `pivot_longer()` a bit limited, usually only reshaping one variable into the long form. In my data and examples for my classes, I often have several time-varying variables to reshape, not just 1 or 2. 

Below, I `select()` the variables that I want to work with, and use `pivot_longer()`. 
I can specify the non-time varying variables in the `cols()` argument, then the default machinery works pretty well, **as long as my variable names are nice and have names like** **_x_1_**. 

```{r}
library(tidyr)
e.long.comp<-eclsk.sub%>%
  rename(wt = w4c4p_40,strata= w4c4p_4str, psu = w4c4p_4psu)%>%
  select(childid,male, hisp, black, asian, nahn, other,wt, strata, psu, #time constant
         height_1, height_2, height_4, #t-varying variables
         ageyrs_1, ageyrs_2, ageyrs_4,
         chhealth_1, chhealth_2, chhealth_4,
         foodinsec_1, foodinsec_2, foodinsec_4,
         hhses_1, hhses_2, hhses_4,
         math_1,math_2, math_4)%>%
  pivot_longer(cols = c(-childid, -male, -hisp, -black, -asian,-nahn, -other, -wt, -strata, -psu), #time constant variables go here
               names_to  = c(".value", "wave"), #make wave variable and put t-v vars into columns
               names_sep = "_")%>% #all t-v variables have _ between name and time, like age_1, age_2
  filter(complete.cases(.))%>%
  arrange(childid, wave)


head(e.long.comp)
```


### Visualization of longitudinal data

```{r}
library(ggplot2)

first10<-unique(e.long.comp$childid)[1:10]

sub<-e.long.comp%>%
  filter(childid%in%first10)

ggplot(sub, aes(x=ageyrs, y=math))+geom_point()+ geom_smooth(method='lm',formula=y~x)+facet_wrap(~childid,nrow = 3)+ggtitle(label = "Change in Math score across age", subtitle = "First 10 children in ECLS-K 2011")

```