---
title: "Mapping DHS Survey Estimates to ADM Level 2 Geographies"
description: |
  In this brief example, I show how to use the GPS coordinates from the DHS data and merge them to the ADM2 subnational geographic level for the country of Ethopia.
author:
  - name: Corey S. Sparks, Ph.D.
    url: https://github.com/coreysparks
date: 04-15-2021
output:
  distill::distill_article:
    self_contained: false
---


In this brief example, I show how to use the GPS coordinates from the [Demographic and Health Survey data](https://dhsprogram.com/) and merge them to the ADM2 subnational geographic level for the country of Ethiopia. Then I produce estimates using the DHS data for ADM 2 regions of the country.

This is possible by using the GIS capacity of the `sf` package to spatially intersect the DHS points and the ADM 2 polygons. 

```{r}
library(sf)
library(mapview)
```

## Read in dhs points
```{r}
ethpoly <- st_read(dsn = "~/OneDrive - University of Texas at San Antonio//students/fikre/spatial_epi/ETH_adm2.shp")

ethpoly$struct <- 1:dim(ethpoly)[1]

plot(ethpoly["struct"])

```


## Read in dhs sample locations and ADM 2 regions.

The adm2 shapefile can be found in the [Diva GIS international data repository](https://www.diva-gis.org/gdata), or from the [IPUMS International site](https://international.ipums.org/international/gis_harmonized_2nd.shtml) below I use the ADM2 level of administrative geography. 

These locations are not identified in the DHS, but by performing a spatial intersection, we can merge the DHS survey locations to the ADM 2 units



```{r}
eth_dots<-st_read("~/OneDrive - University of Texas at San Antonio//students//fikre/ethiopia_dhs/ETGE52FL/ETGE52FL.shp")
eth_dots <- eth_dots[eth_dots$LATNUM>0,]
```

```{r}
eth_adm2<-st_read("~/OneDrive - University of Texas at San Antonio//students//fikre/spatial_epi/ETH_adm2.shp")
```

```{r}
#merge dots to administrative data
eth_dots2000<-st_intersection(eth_dots, eth_adm2)

mapview(eth_dots["DHSCLUST"])+mapview(eth_adm2["NAME_2"])
```

## Read in DHS survey and recode stunting outcome

Next, I use the DHS survey data to estimate the prevalence of stunting in the ADM 2 regions. 

Here I use all children in the *household recode file*. This allows for the inclusion of all children in households, not just the ones born to respondents of the woman's individual survey. If we wanted to use other characteristics of the mother of a particular child in our analysis, then I would use the *childrens' recode file* instead. Since this post is only concerned with estimating raw prevalence of stunting, the household file works for my purpose. 

In 2000, the height and weight variables with the WHO standard are in a separate file, so I merge them according to DHS guidelines. 


```{r}
library(haven)
library(dplyr)
dhs2000<-read_dta("~/OneDrive - University of Texas at San Antonio//students//fikre/ethiopia_dhs/ETPR41DT/ETPR41FL.DTA")
dhs2000<-zap_labels(dhs2000)

htwt00<-read_dta("~/OneDrive - University of Texas at San Antonio//students//fikre/ethiopia_dhs/ETHW41DT/ETHW41FL.DTA")
htwt00<-zap_labels(htwt00)

dhs2000<- merge(dhs2000, htwt00, by.x = c("hhid", "hvidx"), by.y = c("hwhhid", "hwline"))

dhs2000<- dhs2000%>%
  filter(hv103==1 & hc1<60)

library(car)
dhs2000$stunting<-ifelse(dhs2000$hc70/100<=-2&dhs2000$hc70/100!=-2,1,0)
#dhs2000$sex<-dhs2000$hc27

library(dplyr)
dhs2000<-dhs2000%>%
  mutate(wt = hv005/1000000)%>%
  filter(complete.cases(stunting))%>%
  select(hv001,stunting, wt, hv021, hv022)
```

## Merge survey data to sample locations

```{r}
dhs2000m<-merge(dhs2000, eth_dots2000, by.x="hv001", by.y="DHSCLUST")

```


## Create survey estimates for new regions after spatial intersection

```{r}
library(survey)
options(survey.lonely.psu = "adjust")
des<-svydesign(ids = ~hv021, strata = ~hv022, weights = ~wt, data=dhs2000m)
names(dhs2000m)
est.stunt <- svyby(~stunting, ~ID_2, des, FUN=svymean, na.rm=T)

head(est.stunt)

```

## merge estimates to map and map stunting prevalence

```{r}
library(tigris)
library(mapview)
mdat<- geo_join(ethpoly, est.stunt, "ID_2","ID_2")
mapview(mdat["stunting"])


```